Joint estimation of phylogenies and ancestral states is not consistent

### 2017-12-14 joint_inf.tex
address (almost all) comments

\newcommand{\alphabet}{\mathcal{A}}
\newcommand{\fullAlignment}{\mathbf{Y}}
\newcommand{\alignmentColumn}{\mathbf{y}}
\newcommand{\alignmentColumnRV}{Y}
\newcommand{\siteSplit}{\tilde{s}}
\newcommand{\siteSplitSet}{\mathcal{S}}
\newcommand{\fullAncestralStates}{\mathbf{H}}
\newcommand{\ancestralStateColumn}{\mathbf{h}}
\newcommand{\ancestralStateColumnRV}{H}
\newcommand{\ancestralSplit}{\tilde{h}}
\newcommand{\ancestralSplitSet}{\mathcal{H}}
\newcommand{\ancestralSplitPartition}{\eta}
\newcommand{\fullAncestralSplitPartitions}{\boldsymbol\eta}

@@

Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and proceed to estimate the internal branch parameter $w$.
\begin{restatable}{theorem}{restrictedBranchInconsist}
\label{thm:restricted-bl}
Let
$$
\alpha_1 := \alpha_1(x^*, y^*) = \frac{1}{8} \left(1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\alpha_2 := \alpha_2(x^*, y^*) = \frac{1}{8}\left(1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
The maximum likelihood value $\hat{w}$ is equal to 1 if
$$
\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
$$
\end{restatable}
This theorem allows us to demarcate a region of bad internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch, and then a branch length is estimated to be zero (i.e.\ have fidelity $\hat{w} = 1$).
As an intuition for the theoretical development, seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.

@@

The general case is more challenging to analyze and so we will get somewhat weaker bounds.
In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
%EM This doesn't look like the same inequality to me.
Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
We show that there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
We fix bounds on these estimates and show that, in the general case, joint estimation either estimates $\hat{w}$ to be 1 or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
\begin{restatable}{theorem}{generalBranchInconsist}
\label{thm:general-bl}
Define $\gamma(x, y) = 4xy$.
For
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*),
$$
bounds
$$
\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
$$
$$
\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
$$
and
$$
\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}),
$$
the maximum likelihood value $\hat{w}=1$ when
$$
-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
$$
\end{restatable}
This theorem can be used to show incorrect branch parameter estimates as follows.
If we don't want to allow any error in pendant branches, we can use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section (Fig.~\ref{fig:bl-inconsistency}).
For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
$$
x^*-x_{L} \le \hat{x} \le x^*+x_{U}
$$
and
$$
y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
$$
Then useful $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
In this case, joint inference will either estimate $\hat{w}$ to have fidelity 1 or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.

@@

We provide a full description of our optimization procedure in the Appendix, but briefly, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher likelihood.
We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$.
We do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
%EM Need GH link here.

@@

While we have shown inconsistency in both topology and branch parameter estimation, there is substantial scope for future work to make these results more precise and more general.
The techniques used to obtain upper and lower bounds for the likelihoods in the topology estimation case provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
Similarly, for the general case in estimating branch lengths, we were only able to provide a conservative region of overestimation, and the unusual shape we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) begs further investigation.
Empirical validation shows that the general case is not unlike the restricted case.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
%EM Any intuition of if the effect will go away on larger trees? What about more complex models?
%EM Also, is there any hope that an alternate profile likelihood formulation could lead to more consistent estimation? It'd be nice to end with something like that.

>>>>

\newcommand{\alphabet}{\mathcal{A}}
\newcommand{\fullAlignment}{\mathbf{Y}}
\newcommand{\alignmentColumn}{\mathbf{y}}
\newcommand{\alignmentColumnRV}{Y}
\newcommand{\siteSplit}{\tilde{y}}
\newcommand{\siteSplitSet}{\mathcal{Y}}
\newcommand{\fullAncestralStates}{\mathbf{H}}
\newcommand{\ancestralStateColumn}{\mathbf{h}}
\newcommand{\ancestralStateColumnRV}{H}
\newcommand{\ancestralSplit}{\tilde{h}}
\newcommand{\ancestralSplitSet}{\mathcal{H}}
\newcommand{\ancestralSplitPartition}{\eta}
\newcommand{\fullAncestralSplitPartitions}{\boldsymbol\eta}

@@

Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and proceed to estimate the internal branch parameter $w$.
\begin{restatable}{theorem}{restrictedBranchInconsist}
\label{thm:restricted-bl}
Let
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
The maximum likelihood value $\hat{w}$ is equal to 1 if
$$
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0.
$$
\end{restatable}
This theorem allows us to demarcate a region of bad internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch, and then a branch length is estimated to be zero (i.e.\ have fidelity $\hat{w} = 1$).
As an intuition for the theoretical development, seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.

@@

The general case is more challenging to analyze and so we will get somewhat weaker bounds.
In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
%EM This doesn't look like the same inequality to me.
%das: updated
Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
We show that there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
We fix bounds on these estimates and show that, in the general case, joint estimation either estimates $\hat{w}$ to be 1 or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
\begin{restatable}{theorem}{generalBranchInconsist}
\label{thm:general-bl}
Define $\gamma(x, y) = 4xy$.
For
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*),
$$
bounds
$$
\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
$$
$$
\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
$$
and
$$
\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}),
$$
the maximum likelihood value $\hat{w}=1$ when
$$
-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
$$
\end{restatable}
This theorem can be used to show incorrect branch parameter estimates as follows.
If we don't want to allow any error in pendant branches, we can use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section (Fig.~\ref{fig:bl-inconsistency}).
For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
$$
x^*-x_{L} \le \hat{x} \le x^*+x_{U}
$$
and
$$
y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
$$
Then useful $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
In this case, joint inference will either estimate $\hat{w}$ to have fidelity 1 or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.

@@

We provide a full description of our optimization procedure in the Appendix, but briefly, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher likelihood.
We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$.
We do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies\footnote{\url{https://github.com/matsengrp/joint-inf/blob/master/joint_inf_plot.py}}.
%EM Need GH link here.

@@

While we have shown inconsistency in both topology and branch parameter estimation, there is substantial scope for future work to make these results more precise and more general.
The techniques used to obtain upper and lower bounds for the likelihoods in the topology estimation case provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
Similarly, for the general case in estimating branch lengths, we were only able to provide a conservative region of overestimation, and the unusual shape we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) begs further investigation.
Empirical validation shows that the general case is not unlike the restricted case.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
%EM Any intuition of if the effect will go away on larger trees? What about more complex models?
%EM Also, is there any hope that an alternate profile likelihood formulation could lead to more consistent estimation? It'd be nice to end with something like that.
%das: wow, that would be an undertaking. The only thing I know off the top of my head are things like sieves, which I recall Paul Edlefsen being jazzed about but don't know if he uses them in this way. I can think more but it might just be a vague throwaway line.
%das:todo
